An Effective Model of HL-60 Differentiation

### 2017-04-26 Paper_v2.tex
Working on discussion
Working on text

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {An Effective Model of the Retinoic Acid Induced HL-60 Differentiation Program}}}
\vspace{0.05in}
{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Wei Dai, Katherine Rogers, Adithya Sagar, Rodica P. Bunaciu$^{\dag}$, Johanna Congleton$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering} and {$^{\dag}$Department of Biomedical Sciences, {Cornell University, Ithaca NY 14853}}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective modeling of HL-60 differentiation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Frontiers~in~Systems~Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak

@@

The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters which appeared in the control laws regulating these programs
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the p21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted p21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and wash robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.

@@

The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
All simulations were conducted using the best fit parameter set.
The first four modes described $>$99\% of the gene knockout variance,
with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
However, this analysis did not give insight into which upstream components of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

Man-made engineered systems, or naturally occurring cell fate decision programs incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
The current model contained the Gfi-1 and Egr-1/2 agonistic switch.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while reinforcing feedback is often undesirable in human engineered systems, it is at the core of a diverse variety of
cell fate programs and other networks important to human health.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {An Effective Model of the Retinoic Acid Induced HL-60 Differentiation Program}}}
\vspace{0.05in}
{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Johanna Congleton$^{\dag}$, Wei Dai, Katherine Rogers, Adithya Sagar, Rodica P. Bunaciu$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering} and {$^{\dag}$Department of Biomedical Sciences, {Cornell University, Ithaca NY 14853}}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective modeling of HL-60 differentiation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{?}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak

@@

The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters for the signal initiation, and phenotype modules
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the p21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted p21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and was robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.

@@

The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
We conducted pairwise knockout simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
A panel of twenty parameter sets was sampled, and the average displacement matrix was decomposed.
The first four modes described $>$99\% of the gene knockout variance,
with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
However, this analysis did not give insight into which upstream components of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

Naturally occurring cell fate decisions often incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
The current model contained the Gfi-1 and Egr-1/2 agonistic switch; however, its significance was unclear for HL-60 cells.
The expression of Gfi-1, Egr-1/2, C/EBPa and PU.1 was not consistent with the canonical lineage pattern expected from literature.
For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
remained constant following ATRA exposure.
Literature evidence in nonmalignant myelomonocytic fate selection has shown that Gfi-1 and EGR-1/2 promote granulocytic and monocytic differentiation, respectively \cite{Laslo2006}.
Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while reinforcing feedback is often undesirable in human engineered systems, it is at the core of a diverse variety of
cell fate programs and other networks important to human health.
